### 2020-03-02 manuscript.tex
AVMVN

In Bayesian inference, the way parameters are represented in the model can affect the mixing ability of the model and the meaning of the model itself. Three methods for parameterising substitution rates are described below, and are later evaluated in \textbf{\nameref{sect:results}}. Each parameterisation technique is associated with i) an abstraction of the rates $\vec{\mathcal{R}}^{\,}$, ii) some function for transforming this parameter into real rates $r(\vec{\mathcal{R}}^{\,})$, and iii) a prior density function of the abstraction $p(\vec{\mathcal{R}}^{\,} | \sigma) $. The three methods are summarised in \textbf{Fig \ref{fig:rateparams}}.

@@

@@

\textbf{2. Time to convergence}. Two independent MCMC chains are run and the time is measured until: a) the absolute difference in clade posterior probability between the two chains is less than 0.05 for all clades, b) the Rubin-Gelman statistic $\hat{R}$ \cite{gelman1992inference} of every estimated parameter is less than 1.05, and c) the effective sample size \cite{rambaut2018posterior} of every estimated parameter is greater than 100 in each chain. \\

@@

For the latter two criteria, methodologies are benchmarked using the empirical datasets compiled and partitioned \cite{lanfear2016partitionfinder} by Lanfear \cite{lanfear2019Github}. Four datasets (the spider data by Richart et al. 2015 \cite{Richart_2016}; the turtle data by Crawford et al. 2012 \cite{Crawford_2012}; the caterpillar dataset by Kawahara et al. 2013 \cite{Kawahara_2013}; and the songbird data by Moyle et al. 2016 \cite{Moyle_2016}) are used during the optimisation stage, where each setting is run $10 \times$ and the average statistics across the $10 \times 2 = 20$ chains are reported. All 25 datasets presented in \textbf{Table \ref{table:datasets}} are benchmarked together $1 \times$ each at the end of the protocol. Datasets and BEAST2 .xml templates are provided in the GitHub repository accompanying this article.

@@

% https://github.com/roblanf/BenchmarkAlignments
\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l l l|} 
 \hline
  & $N$ & $P$ & $L$ (kb) & $L_\text{eff}$ (kb) & Ref. \\

\textbf{1}  &  \textbf{6}  &  1/4/8/\textbf{16}  &  0.4/2.1/10.2/\textbf{13.7}  &  0.1/0.5/1/\textbf{1.1}  &  \textbf{Richart 2015} \cite{Richart_2016} \\

\textbf{2}  &  \textbf{10}  &  1/4/8/\textbf{16}  &  0.4/1.5/3.1/\textbf{6.7}  &  0.1/0.3/0.5/\textbf{0.9}  &  \textbf{Crawford 2012} \cite{Crawford_2012} \\

3  &  11  &  1/4/8/16  &  0.5/2.2/4.9/10.1  &  0.1/0.4/0.7/1.3  &  Leache 2015 \cite{Leache_2015} \\

4  &  18  &  1/4/8/16  &  0.4/1.8/2.9/6.7  &  0.1/0.3/0.4/0.9  &  Meiklejohn 2016 \cite{Meiklejohn_2016} \\

5  &  27  &  1/4/8/16  &  0.8/1.4/2/4.3  &  0.3/0.6/1/1.7  &  Faircloth 2013 \cite{Faircloth_2013} \\

6  &  33  &  1/4/8/16  &  0.4/1.7/2.6/5.1  &  0.1/0.5/0.7/1.5  &  McCormack 2013 \cite{McCormack_2013} \\

7  &  38  &  1/4  &  0.4/2.1  &  0.2/0.6  &  Bergsten 2013 \cite{Bergsten_2013} \\

8  &  38  &  1/4/8/16  &  1/2.4/8.1/14.9  &  0.7/1.8/5.6/10.1  &  Ran 2018 \cite{Ran_2018} \\

9  &  41  &  1/3  &  0.4/1.7  &  0.3/1.1  &  Brown 2012 \cite{Brown_2012} \\

10  &  44  &  1/3  &  0.6/1.9  &  0.2/0.8  &  Cognato 2001 \cite{Cognato_2001} \\

11  &  44  &  1/4/7  &  0.8/2.9/5.9  &  0.3/0.8/1.8  &  Dornburg 2012 \cite{Dornburg_2012} \\

12  &  51  &  1/4/6  &  0.6/3.5/5.4  &  0.1/0.9/1.8  &  Sauquet 2011 \cite{Sauquet_2011} \\

13  &  61  &  1/4/8/16  &  0.9/4.2/7.5/15  &  0.7/2.8/4.6/9.6  &  Broughton 2013 \cite{Broughton_2013} \\

14  &  69  &  1/2  &  0.7/0.8  &  0.1/0.1  &  Devitt 2013 \cite{Devitt_2013} \\

\textbf{15}  &  \textbf{70}  &  1/\textbf{3}  &  0.7/\textbf{2.2}  &  0.3/\textbf{0.9}  &  \textbf{Kawahara 2013} \cite{Kawahara_2013} \\

16  &  78  &  1/4/8/16  &  0.4/1.9/4.1/7.2  &  0.3/1.7/3.6/6.2  &  Cannon 2016 \cite{Cannon_2016} \\

17  &  79  &  1/4/8/16  &  0.1/1.3/0.8/4.5  &  0/0.1/0.1/0.8  &  Oaks 2011 \cite{Oaks_2011} \\

18  &  94  &  1/4/8/11  &  0.1/1.8/3/3.7  &  0.1/0.8/1.3/1.7  &  Rightmyer 2013 \cite{Rightmyer_2013} \\

\textbf{19}  &  \textbf{106}  &  \textbf{1}/4/8/16  &  \textbf{0.6}/3.1/5.1/11.6  &  \textbf{0.3}/1.2/2/4.5  &  \textbf{Moyle 2016} \cite{Moyle_2016} \\

20  &  110  &  1/4/8/16  &  0.5/2.1/3.8/7.1  &  0.3/1.4/2.5/4.3  &  Fong 2012 \cite{Fong_2012} \\

21  &  152  &  1/4/5  &  1/3.5/3.6  &  0.6/1.6/1.6  &  Day 2013 \cite{Day_2013} \\

22  &  187  &  1/4/8/16  &  0.3/0.9/1.4/3.5  &  0.2/0.7/1.1/2.9  &  Branstetter 2017 \cite{Branstetter_2017} \\

23  &  197  &  1/4/8/14  &  0.9/3.1/6.1/11.6  &  0.3/1.6/3.5/6.3  &  Horn 2014 \cite{Horn_2014} \\

24  &  235  &  1/4/8/16  &  2.3/6.4/8.5/25.7  &  0.9/2.3/3.2/12.2  &  Reddy 2017 \cite{Reddy_2017} \\

25  &  237  &  1/4/5  &  0.8/2.4/3.1  &  0.2/1.2/1.5  &  Murray 2013 \cite{Murray_2013} \\

 \hline
\end{tabular}
\caption{Empirical datasets used during benchmarking, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are also specified. Where a dataset is sampled more than once, the dimensions of its multiple samples are separated by `/'. The datasets in bold (Richart 2015 \cite{Richart_2016}, Crawford \cite{Crawford_2012}, Kawahara \cite{Kawahara_2013}, and Moyle \cite{Moyle_2016}) are used during the optimisation protocol. }
\label{table:datasets}
\end{table}

@@

We compared the Narrrow Exchange Rate (NER) operators described in \textbf{\nameref{sect:NER}}. This protocol selects the best among 48 NER (no random walk) and 48 NERw ($\text{Bactrian}(0.95)$ random walk) operators, and has two phases. First, the best of the 96 is selected by comparing operator acceptance rates on simulated data. Second, the selected operator is benchmarked with respect to convergence time and sampling rate on real data (\textbf{Table \ref{table:datasets}}). All the analyses in this section invoke the \textit{quant} parameterisation and $\text{Bactrian}(0.95)$ proposal kernels on clock model parameters.

>>>>

In Bayesian inference, the way parameters are represented in the model can affect the mixing ability of the model and the meaning of the model itself \cite{gelman2004parameterization}. Three methods for parameterising substitution rates are described below, and are later evaluated in \textbf{\nameref{sect:results}}. Each parameterisation technique is associated with i) an abstraction of the rates $\vec{\mathcal{R}}^{\,}$, ii) some function for transforming this parameter into real rates $r(\vec{\mathcal{R}}^{\,})$, and iii) a prior density function of the abstraction $p(\vec{\mathcal{R}}^{\,} | \sigma) $. The three methods are summarised in \textbf{Fig \ref{fig:rateparams}}.

@@

\subsection*{A guided adaptive leaf rate operator}

A \textit{guided} operator incorporates knowledge about neighbouring states, while an \textit{adaptive} operator undergoes a training process to improve its efficiency over time \cite{roberts2007coupling}. In previous work, parsimony scores and conditional clade probabilities of neighbouring trees have been employed by guided tree operators \cite{hohna2012guided,zhang2019using,meyer2019adaptive} and the latter has also been explored as the basis of  adaptive tree operators \cite{hohna2012guided,meyer2019adaptive}. The (adaptive) mirror kernel \cite{thawornwattana2018designing} learns a target distribution which acts as a `mirror image' of the current point $x$.  The adaptable variance multivariate normal (AVMVN) kernel \cite{baele2017adaptive,suchard2018bayesian} learns correlations between parameters during MCMC. Baele et al. 2017  observed a large increase ($\approx 5-10 \times$) in sampling efficiency from using the AVMVN kernel on clock rates and substitution model parameters across partitions \cite{baele2017adaptive}.

In this article we consider application of the AVMVN kernel to the branch rates of leaf nodes. This operator is not readily applicable to internal node branch rates due to their dependency on the topology of the tree.

\subsubsection*{AVMVN kernel}
\label{AVMVN_sect}

The AVMVN kernel assumes its parameters live in $x \in \mathbb{R}^N$ and that these parameters follow a multivariate normal distribution with covariance matrix $\Sigma_N$. Hence, the kernel operates on the logarithmic or logistic transformation of the $N$ leaf branch rates, depending on the rate parameterisation:

\begin{align}
	x_i = \begin{cases} \log r_i \text{ for } \textit{real} \\
						\log \frac{q_i}{1 - q_i} \text{ for } \textit{quant}  \end{cases}
\end{align}

where $r_i$ is a real rate and $q_i$ is a rate quantile. The AVMVN probability density is defined by

\begin{align}
	\mathcal{AVMVN}(x) = (1-\beta) \mathcal{MVN}\big(x, \frac{\Sigma_N}{N} \big) + \beta \mathcal{MVN}\big(x, \frac{\mathbb{I}_N}{N} \big),
\end{align}

where $\mathcal{MVN}$ is the multivariate normal probability density. $\beta \; (= 0.05)$ is a constant which determines the fraction of the proposal determined by the identity matrix ${I}_N$, as opposed to the covariance matrix $\Sigma_D$ which is trained during MCMC.

The AVMVN proposal kernel is computed as

\begin{align}
	& x^\prime \leftarrow x + \sum\limits_{i=1}^N \sum\limits_{j=i}^N c_{i,j} \times s\Sigma \\
	\text {where } & c = \texttt{cholesky} \{ (1-\beta) \frac{\Sigma_N}{N} + \beta \frac{\mathbb{I}_N}{N} \}.
\end{align}

The \texttt{cholesky}$\{Y\}$ decomposition returns a lower diagonal matrix $L$, with positive real diagonal entries, such that $Y = LL^\prime$ \cite{lindstrom1988newton, pourahmadi2007cholesky}. $s$ is a tunable step size parameter and $\Sigma$ is a random variable drawn from a proposal kernel (uniform or Bactrian for instance). Our BEAST2 implementation of the AVMVN kernel is adapted from that of BEAST \cite{suchard2018bayesian}.

In \nameref{sect:results}, we evaluate this operator for its ability to estimate leaf rates. As the size of the covariance matrix $\Sigma_N$ grows with the number of taxa $N$, AVMVN is hypothesised to work well on small trees but become less efficient with larger taxon sets.

@@

\textbf{2. Time to convergence}. Two independent MCMC chains are run and the time is measured until: a) the absolute difference in clade posterior probability between the two chains is less than 0.05 for all clades, b) the Rubin-Gelman statistic $\hat{R}$ \cite{gelman1992inference} of every estimated parameter is less than 1.05, and c) the effective sample size \cite{rambaut2018posterior} of every estimated parameter is greater than 200 in each chain. \\

@@

For the latter two criteria, methodologies are benchmarked using both simulated and empirical datasets -- the latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}). Each setting is run $10 \times$ per dataset and the average statistics across the $10 \times 2 = 20$ chains are reported. .

%All 25 datasets presented in \textbf{Table \ref{table:datasets}} are benchmarked together $1 \times$ each at the end of the protocol. Datasets and BEAST2 .xml templates are provided in the GitHub repository accompanying this article.

@@

% https://github.com/roblanf/BenchmarkAlignments
\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l l l|} 
 \hline
  & $N$ & $P$ & $L$ (kb) & $L_\text{eff}$ (kb) & \textbf{Description} \\

1  &  6  &  16  &  13.7  &  1.1  &  Spiders (Richart 2015 \cite{Richart_2016}) \\

2  &  18  &  16  &  6.7  &  0.9  &  Gallopheasants (Meiklejohn 2016 \cite{Meiklejohn_2016}) \\

3  &  33  &  16  &  5.1  &  1.5  &  Birds (McCormack 2013 \cite{McCormack_2013}) \\

4  &  40  &  1 &  3.0  &  0.6-1.8  &  \textit{Simulated data} $\times 10$ \\

5  &  44  &  3  &  1.9  &  0.8  &  Bark beetles (Cognato 2001 \cite{Cognato_2001}) \\

6  &  51  &  6  &  5.4  &  1.8  &  Southern beeches (Sauquet 2011 \cite{Sauquet_2011}) \\

7  &  70  &  3  &  2.2  &  0.9  &  Caterpillars (Kawahara 2013 \cite{Kawahara_2013}) \\

8  &  94  &  4  &  1.8  &  0.8  &  Bees (Rightmyer 2013 \cite{Rightmyer_2013}) \\

9  &  106  &  1  &  0.6  &  0.3  & Songbirds (Moyle 2016 \cite{Moyle_2016}) \\

10  &  110  &  1  &  0.5  &  0.3  &  Vertebrates (Fong 2012 \cite{Fong_2012}) \\

 \hline
\end{tabular}
\caption{Datasets used during benchmarking, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of patterns $L_\text{eff}$ are also specified. The 9 empirical datasets are benchmarked $10 \times$ per setting, whereas each of the 10 simulated datasets are benchmarked once per setting.}
\label{table:datasets}
\end{table}

@@

The \textbf{\nameref{sect:NER}} (NER) operators are evaluated. This protocol selects the best among 48 NER (no random walk) and 48 NERw ($\text{Bactrian}(0.95)$ random walk) operators, and has two phases. First, the best of the 96 is selected by comparing operator acceptance rates on simulated data. Second, the selected operator is benchmarked with respect to convergence time and sampling rate on real data (\textbf{Table \ref{table:datasets}}). The analyses in this section invoke the \textit{quant} parameterisation and $\text{Bactrian}(0.95)$ proposal kernels on clock model parameters.
